---
title: "Predicting Grades - Part 01"
author: "Caynan Sousa"
date: "3/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(recommenderlab)
library(DMwR)
library(dplyr)
library(reshape2)
library(ggplot2)
library(Matrix)
```

# 1. Introduction
Let's practice the use of recommendation systems.

We want to predict the grade a student will obtain on the 4th semester of the Computer Science course at UFCG (My university), for this let's use collaborative filtering.

The basic idea is to use collaborative filtering on the users for:  
  - Find the most similar students, using the classes taken and the grades obtained to the targeted students.  
  - Based on those neighbors predict the grade our student might obtain in the 4th semester classes. 

### First things first

Let's get our data that is hosted in our repo on github.

```{r load_data}
COL_NAMES <- c("student_id", "year", "term", "semester_coursed", "course_id", "course_name", "course_semester", "grade")

train <- read.csv("https://raw.githubusercontent.com/caynan/data_analysis_2/master/04_atividade/data/lab4_part1_train_data.csv")
train <- na.omit(train)
colnames(train) <- COL_NAMES
# Creating factors
train$student_id <- as.factor(train$student_id)
train$course_id <- as.factor(train$course_id)

test <- read.csv("https://raw.githubusercontent.com/caynan/data_analysis_2/master/04_atividade/data/lab4_part1_test_data.csv")
test <- na.omit(test)
colnames(test) <- COL_NAMES
```

# 2. Data Manipulation

Let's transform our dataset to be wide using reshape and remove some columns we're not going to be using on our collaborative recommender. We're also, converting our data into realRatingMatrix data structure which is used by the recommenderlab library.

```{r widen_data}
# ME, PLP, Logica, OAC, ES, SI1, LOAC
COL_4_TERM <- c("1114114", "1411169", "1411173", "1411175", "1411177", "1411178", "1411182")

widen_data <- function(df) {
  data.reshaped.by.course <- df %>%
    group_by(student_id, course_id) %>%
    mutate(grade = max(grade)) %>%
    ungroup() %>%
    distinct(student_id, course_id, .keep_all = TRUE) %>%
    select(student_id, course_id, grade) %>%
    dcast(student_id ~ course_id, value.var = "grade") #, fill=0
  
  row.names(data.reshaped.by.course) <- data.reshaped.by.course$student_id
  data.reshaped.by.course$student_id <- NULL
  
  return(data.reshaped.by.course)
}

toRatingMatrix <- function(df) {
  temp <- as.matrix(df)
  temp <- as(temp, "realRatingMatrix") 
  
  return(temp)
}

train_widen <- widen_data(train)
train_rating_matrix <- toRatingMatrix(train_widen)
```

Let's prepare our data to be used by the `Recommender` method.

```{r removing_columns}
# Grades of 4th semester classes should be erased
test_widen <- widen_data(test) # we're using this at the end to evaluate our model.
test_cleaned <- test_widen
for(i in 1:length(COL_4_TERM)) {
  # We set all the columns we want to make predictions to NA.
  test_cleaned[COL_4_TERM[i]] <- NA
}

# Create rating matrix
test_rating_matrix <- toRatingMatrix(test_cleaned)
```

# 3. Recommender System

Let's train our recommender system using a User Based approach (UBCF).

```{r training_recommender}
# Building model
model <- Recommender(train_rating_matrix, method = "UBCF", 
                     param=list(normalize = "center", method="pearson", nn=10))
```

And now let's use the model we just created to predict the grades of the students on our test set.

```{r predicting_grades}
# Making predictions 
prediction <- predict(model, test_rating_matrix, type="ratings")

# Cleanning Predicted Data and making it wide
t <- as(prediction, "data.frame")
colnames(t) <- c("student_id", "course_id", "grade")
predicted_test <- widen_data(t)
```


Calculating the RMSE for our prediction.

```{r rmse}
# Get only 4th semester courses for predicted and real values
predicted_test <- select(predicted_test, one_of(COL_4_TERM))
real_test <- select(test_widen, one_of(COL_4_TERM))

# Calculate RMSE
rmse <- sqrt( mean( (predicted_test - real_test)^2 , na.rm = TRUE ) )
rmse
```

## 3.1 Imputing Data

If we stop to think about it, we're [imputing missing data][imputation wiki]. So we can adapt the `knnImputation` method from the `DMwR` package, which is used to fill NAs in a data frame using the weigthed average of their K-nearest neighbors. Which reduce our work to:

```{r using_DMwR}
DMwR_prediction <- knnImputation(test_cleaned, scale=F, k = 10, distData = train_widen)
```

The only caveat is that we're using euclidean distance here in contrast with the previous way where we used Pearson to calculate the distance between students.

Let's see if we obtained a better RMSE using this method

```{r rmse_DMwR}
DMwR_prediction <- select(DMwR_prediction, one_of(COL_4_TERM))
rmse_DMwR <- sqrt( mean( (DMwR_prediction - real_test)^2 , na.rm = TRUE ) )
rmse_DMwR
```

We obtained slightly worse results, which I theorize is given the fact we're using Euclidean Distance, given that Pearson find the students on which their grades vary in the same way and that Euclidean distance find the students with the closest grades to one another.

## Do we have any missing predictions?

The percentage of missing predictions is zero for both methods, we have at least one other student that is similar to the ones we're trying to predict their grade, which given that we have 2107 students in our training set is an expected result.

### That is it.

[imputation wiki]: https://en.wikipedia.org/wiki/Imputation_(statistics)