---
title: "predicting_evasion"
author: "Caynan Sousa"
date: "26/2/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# For data manipulation and tidying
library(dplyr)
library(tidyr)
library(Hmisc)
library(reshape2)

# For data visualizations
library(ggplot2)
library(highcharter)
library(igraph)
# library(networkD3)
library(htmlwidgets)

# For machine learning
library(caret)
library(ranger)
library(e1071)
```

## Introduction


### Importing Dataset

```{r load_data}
# Load data and set it to `data`
data.url <- "https://raw.githubusercontent.com/caynan/data_analysis_2/master/03_atividade/treino_classificacao_v2.csv"
data <- read.csv(data.url)

# rename columns
colnames(data) <- c("student_id", "course_id", "course_name", "year", "term", "mean", "evaded")
```


### Data Structure and Variables

Great! Now that our data is imported, we can get a look at it. I'll start by looking at it's structure.

```{r data_structure}
str(data)
```

So as we expected we're working with a data frame of 7 variables and 7874 observations.

Our 7 variables are:   
- **student_id**: 
- **course_id**:
- **course_name**:
- **year**:
- **term**:
- **mean**:
- **evaded**: 

I'm going to change the data type of the student_id, course_id, term and year columns. Apart from year, which is ordinal, all the other should be factors (nominal). 

```{r change_datatypes}
# change data types for student_id, course_id, evaded, year and term
data$student_id <- as.factor(data$student_id)
data$course_id <- as.factor(data$course_id)
data$term <- as.factor(data$term)
data$year <- as.ordered(data$year)
```

A few basic questions that we might have about our data is:

How many years are we looking here?
```{r how_many_years}
levels(data$year)
```

Cool, so we're dealing with data since the year 2000 up to 2015.

How many students data do we have in our data?
```{r how_many_students}
data %>% summarise(n_distinct(student_id))
```
only **1351** students were accepted into Computer Science in the past 15 years, wow!

## Can we predict if a student will evade after the first semester?
bla bla bla, reasons why this is important

### Feature Engineering

I'm going to use some simple machine learning models (logistic regression and decision trees) to see if I can predict if a student will evade or not after the first
semester. Summarising, we're trying to predict if a student evaded or not (dependent variable)  and for that we're going to use 10 (+ 1 extra) independent variables:  
 - Grade on Calculus I (aka: calculus)
 - Grade on Vectorial Algebra (aka: vectorial)
 - Grade on Intro to CS (aka: ICS)
 - Grade on Programming I (aka: P1)
 - Grade on Lab. of Programming I (aka: LP1)
 - Grade on Reading and Producing Text (aka: RPT)
 - General GPA
 - Comp. Sci. GPA (taking in consideration only classes from the Systems and Computers Department)
 - Math GPA (taking in consideration only classes from the Math and Statistics Department)
 - Term (if they got accepted into the first or the second semester)

But first, we need to transform our dataset to fit into a frame with the needed variables.

```{r reshaping_data}
# reshape data
# - only one entry per student_id
# - reshape course mean into column

# get only one entry per student
unique.data <- data %>%
  group_by(student_id) %>% select(student_id, evaded, year, term) %>% unique()

# reshape courses into columns
data.reshaped.by.course <- data %>%
  select(student_id, course_name, mean) %>%
#  mutate(mean = ifelse(is.na(mean), 0, mean)) %>%
  mutate(course_name = as.factor(gsub(" ", ".", course_name))) %>%
  dcast(student_id ~ course_name, value.var = "mean")

# merge the two DFs
mlset <- merge(unique.data, data.reshaped.by.course)
# set column names to smaller easier to type names, why you wonder, because I'm lazy ;p
colnames(mlset) <- c("student_id", "evaded","year", "term", "vectorial", "calculus", "ICS", "LP1", "RPT", "P1")
```

As we can see, we have a lot of NA on the courses columns, which mean that those students don't have record for those courses. This phenomenon could be explained by a number of reasons, could be that the student coursed them elsewhere, or it could be that he failed by the number of absences.

Since is a little bit tricky the reason on why a given student don't have records for a given class, we're going to solve that in two steps:  
- First we need to keep track of how many missing records(NA) we have per student (row)
- Second, let's introduce a simple algorithm to fill those columns, if the number of missing records is greater than 3 (half the number of mandatory classes), let's set the mean to zero (0.0), But if is less than or equal to 3 let's set the value to the mean of the respective course for all other students on that given year and term.

```{r dealing_with_na}
# count the number of missing records
mlset$missing_records <- apply(is.na(mlset), 1, sum)

# implement our algorithm for each one of the courses column
mlset <- mlset %>%
  group_by(year, term) %>%
  mutate(calculus = ifelse(is.na(calculus), ifelse(missing_records > 3, 0, median(calculus, na.rm = TRUE)), calculus)) %>%
  mutate(vectorial = ifelse(is.na(vectorial), ifelse(missing_records > 3, 0, median(vectorial, na.rm = TRUE)), vectorial)) %>%
  mutate(ICS = ifelse(is.na(ICS), ifelse(missing_records > 3, 0, median(ICS, na.rm = TRUE)), ICS)) %>%
  mutate(P1 = ifelse(is.na(P1), ifelse(missing_records > 3, 0, median(P1, na.rm = TRUE)), P1)) %>%
  mutate(LP1 = ifelse(is.na(LP1), ifelse(missing_records > 3, 0, median(LP1, na.rm = TRUE)), LP1)) %>%
  mutate(RPT = ifelse(is.na(RPT), ifelse(missing_records > 3, 0, median(RPT, na.rm = TRUE)), RPT))

# We don't have any use for the year column, so let's drop it.
 mlset$year <- NULL
```

### Is our Data Unbalanced?

Let's take a look if our data is unbalanced, have way more students in one category than on another (way more students who didn't evaded than students who evaded).

```{r more_dropouts}
num_dropouts <- mlset %>%
  summarise(num_students = n(),
            num_dropouts = sum(evaded)) %>%
  ungroup() %>%
  summarise(total_students = sum(num_students),
            total_dropouts = sum(num_dropouts))

total_dropouts.melt <- total_dropouts %>% 
  melt(variable.name="data_class", value.name="total_count")
```

We can plot this information to make it easier to see our distribution.

```{r num_dropouts_plot, echo=FALSE}
ggplot(total_dropouts.melt, aes(data_class, total_count)) +
  geom_bar(stat = "identity", position = "dodge", fill="#56B4E9") +
  geom_text(aes(label=sprintf("%d", total_count)), size = 3) +
  # scale_x_continuous(breaks=seq(2000, 2015, 1)) +
  ylab("Number of Students") +
  xlab("Data Classes")
```

As we can see our data is unbalanced, we have way more students that who didn't evaded than ones that have. To be more precise, we have that only ~9.32% of the data is of students that evaded.
But don't worry we will deal with this in a bit, but first let's see how our prediction models work with this unbalanced data, so we can have a baseline to compare with.


### Modeling

Now that the dataframe is set up the way I want (a single row per student and only the variables I'm interested in), I'm going to randomize the dataset and extract a test set.

```{r create_testing_set}
# Set a random seed
set.seed(42)

# Shuffle rows of the data set
n <- nrow(mlset)
shuffled <- mlset[sample(n), ]

# Perform 80/20 split (80% to training set, 30% to test set)
train_indices <- 1:round(0.8 * n)
train <- shuffled[train_indices, ]

test_indices <- (round(0.8 * n) + 1):n
test <- shuffled[test_indices, ]
```


#### Fitting a baseline model
probably something along the line of predicting that the student will not evade (which is good because of the unbalance)


```{r fitting_model}
myControl <- trainControl(method = "cv", number = 10)

reg_model <- train(as.factor(evaded) ~ term + vectorial + calculus + ICS + LP1 + RPT + P1 + missing_records,
                   data = train, tuneLength = 3, method = "glm", family = "binomial", trControl = myControl, na.action = na.omit)

predict_output <- predict(reg_model, newdata = test)
confusionMatrix(predict_output, test$evaded)
```

model.reg.cc <- train(evadiu ~ cra + num_NA + status + num_finais + C1 + AV + 
                      IC + LP1 + LPT + P1 + cra_cc,
                       data=treino,
                       method="glm",
                       family="binomial",
                       na.action = na.omit)

pred.reg.cc <- predict(model.reg.cc, newdata=teste)
treino.pred.reg.cc <- predict(model.reg.cc, newdata=treino)

acc.reg.cc <- confusionMatrix(pred.reg.cc, teste$evadiu)
acc.reg.cc














































